Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-03-23 content.tex
Fix LaTeX
Builds

BEAST [\cite{drummond2007beast}] is a typical Bayesian phylogenetics tool, 
that needs both character data and priors to infer 
a posterior distribution of phylogenies.
Specifically, for the species tree prior - which describes 
the process of diversification - 
BEAST has built-in priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] models.
These simple tree priors are the most commonly used, as they have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2019beast}],
has a package manager, that allows third-party users 
to extend existing functionalities.
For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, 
\cite{laudanno2020sls}], models that treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or as a burst of simultaneous branching
events [\cite{laudanno2018mbd}], or where diversification rate
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
even concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
\fi
Many diversification models (and their associated probability algorithms) have been developed to account for possible mechanisms.
There are models in which diversification rates are time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or not homogeneous across the entire phylogeny [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, 
\cite{laudanno2020sls}], models that treat speciation as a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or as a burst of simultaneous branching
events [\cite{laudanno2018mbd}].
Others account for trait-dependent regimes, in several ways: where diversification rates
depend on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
even concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Despite such a rich theoretical landscape, only a few of these diversification models are available as a BEAST2 package.

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:create_dd_tree},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny <- create_dd_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/example_30/true_tree.png}
  \caption{
    The tree derived from a DD process, 
    as created by listing~\ref{lst:create_dd_tree}.
  }
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/example_30/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

>>>>

BEAST [\cite{drummond2007beast}] is a typical Bayesian phylogenetics tool, 
that needs both character data and priors to infer 
a posterior distribution of phylogenies.
Specifically, for the species tree prior - which describes 
the process of diversification - 
BEAST has built-in priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] models.
These simple tree priors are the most commonly used, as they have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2019beast}],
has a package manager, that allows third-party users 
to extend existing functionalities.
For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, 
\cite{laudanno2020sls}], models that treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or as a burst of simultaneous branching
events [\cite{laudanno2018mbd}], or where diversification rate
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
even concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].

Many diversification models (and their associated probability algorithms) have been developed to account for possible mechanisms.
There are models in which diversification rates are time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or not homogeneous across the entire phylogeny [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, 
\cite{laudanno2020sls}], models that treat speciation as a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or as a burst of simultaneous branching
events [\cite{laudanno2018mbd}].
Others account for trait-dependent regimes, in several ways: where diversification rates
depend on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
even concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Despite such a rich theoretical landscape, only a few of these diversification models are available as a BEAST2 package.

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:create_dd_tree},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny <- create_dd_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/example_30_314/true_tree.png}
  \caption{
    The tree derived from a DD process, 
    as created by listing~\ref{lst:create_dd_tree}.
  }
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}
